Effectiveness of BNT162b2 booster doses in England: a observational study in OpenSAFELY-TPP

### 2022-03-03 manuscript/draft-manuscript.Rmd
lots of updates to manuscript

---
title: "Effectiveness of BNT162b2 booster doses in England: a observational study in OpenSAFELY"
output:
  html_document:
    keep_md: yes
    self_contained: yes
  word_document: default
  bookdown::html_document2:
    number_sections: no
    toc: no
  pdf_document:
    toc: no
bibliography: references.bib
csl: vancouver.csl
link-citations: yes
zotero: yes
header-includes:
- \usepackage{float} 
- \floatplacement{figure}{H}
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \captionsetup[table]{labelformat=empty}
---

@@

flowchart <- read_csv(fs::path(output_dir_os, "flowchart.csv"))
cohortsummary <- read_csv(fs::path(output_dir_os, "cohortsummary.csv"))
matchsummary <- read_csv(fs::path(output_dir_os, "matchsummary.csv"))
matchsummary_treated <- read_csv(fs::path(output_dir_os, "matchsummary_treated.csv")) %>% filter(treated == 1L)
matchsummary_control <- read_csv(fs::path(output_dir_os, "matchsummary_treated.csv")) %>% filter(treated == 0L)
matchflowchart <- read_csv(fs::path(output_dir_os, "matchflowchart.csv"))
matchcoverage <- read_csv(fs::path(output_dir_os, "matchcoverage.csv"))
incidences <- read_csv(fs::path(output_dir_os, "none", "incidence.csv")) %>% mutate(events=events_1+events_0)

@@

  outcomes <- 
    c(
      "postest",
      #"covidemergency",
      "covidadmitted",
      #"covidcc",
      "coviddeath"
    ) %>% 
    set_names(., .)

km_none <- read_csv(fs::path(output_dir_os, "none", "km.csv"))  
km_vax12_type <- read_csv(fs::path(output_dir_os, "vax12_type", "km.csv"))

km <- 
  bind_rows(km_none, km_vax12_type) %>%
  mutate(
    subgroup_descr = fct_inorder(subgroup_descr),
    subgroup_level = fct_inorder(subgroup_level),
    subgroup_level_descr = fct_inorder(fct_explicit_na(subgroup_level_descr, "All")),
    treatment_descr = fct_inorder(treatment_descr),
    outcome_descr = fct_inorder(outcome_descr),
    outcome_descr_wrap = fct_relabel(outcome_descr, ~str_wrap(., width=20)),
    #model_name = fct_inorder(str_wrap(model_name, 15)),
  )
# 
# cmlinc <- read_csv(fs::path(output_dir_os, "cmlinc.csv")) %>%
#   mutate(
#     treatment_descr = fct_inorder(treatment_descr),
#     outcome_descr = fct_inorder(outcome_descr),
#     model_name = fct_inorder(str_wrap(model_name, 15)),
#     inc_CI_to = paste0(label_number(0.1, scale=1000)(1-survival.ul), " to ", label_number(0.1, scale=1000)(1-survival.ll))
#   )

@@

hazards_none <- read_csv(fs::path(output_dir_os, "none", "effects.csv"))
hazards_vax12_type <- read_csv(fs::path(output_dir_os, "vax12_type", "effects.csv"))

hazards <- 
  bind_rows(hazards_none, hazards_vax12_type) %>%
  mutate(

@@

metahazards_none <- read_csv(fs::path(output_dir_os, "none", "metaeffects.csv"))
metahazards_vax12_type <- read_csv(fs::path(output_dir_os, "vax12_type", "metaeffects.csv"))

metahazards <- 
  bind_rows(metahazards_none, metahazards_vax12_type) %>%
  mutate(

@@

@@

With the approval of NHS England we used the OpenSAFELY-TPP database, covering 40% of GP practices in England and linked to national coronavirus surveillance, hospital episodes, and death registry data, to estimate the effectiveness of a booster dose versus two doses only. On each day of the recruitment period (`r format(as.Date(study_dates$studystart_date), "%d %B %Y")` to `r format(as.Date(study_dates$lastvax3_date), "%d %B %Y")`) each eligible person receiving a BNT162b2 booster dose was matched with an unboosted control who had received two vaccine doses. Primary outcomes were documented SARS-CoV-2 infection, COVID-19-related accident and emergency attendance, COVID-19-related hospital admission, and COVID-19-related death occurring within 10 weeks of recruitment. Hazard ratios were estimated using Cox models adjusting for additional covariates that were not used as matching factors. Follow-up spanned a period in which the Delta variant was predominant.

@@

We matched `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="pfizer") %>% pull(n_matched))` BNT162b2 booster recipients to unboosted controls. During days 14-20 after the booster, the estimated hazard ratio (95% CI) comparing a BNT162b2 booster dose to two doses only was `r filter(hazards, subgroup=="none", outcome=="postest", treatment=="pfizer", model==3, term_left==14) %>% pull(ve_display)` for documented SARS-CoV-2 infection, `r filter(hazards, subgroup=="none", outcome=="covidadmitted", treatment=="pfizer", model==3, term_left==14) %>% pull(ve_display)` for COVID-19 hospital admission, and `r filter(hazards, subgroup=="none", outcome=="coviddeath", treatment=="pfizer", model==3, term_left==14) %>% pull(ve_display)` for COVID-19 death.

@@

<!-- *Participants* `r label_number(1, big.mark=",")(filter(flowchart, crit=="c4") %>% pull(n))` people with a complete primary COVID-19 vaccine course, registered with a GP practice using the TPP SystmOne clinical information system in England. -->

@@

<!-- * full description of booster eligiblity over time -->

<!-- * figure 1 chart with SMD / %-point differences -->

@@

The national COVID-19 vaccination programme in England began a new phase in September 2021 with administration of the first booster COVID-19 vaccine doses in those who had already received their primary vaccination course (a completed first and second dose) (ref <https://www.england.nhs.uk/2021/09/nhs-begins-covid-19-booster-vaccination-campaign/>). Initially offered to those at highest risk of severe disease, eligibility was progressively extended and by 15 December 2021 every adult in England was eligible (<https://www.england.nhs.uk/2021/12/nhs-booster-bookings-open-to-every-eligible-adult/>). Booster doses were initially available no earlier than six months after dose two, but this was reduced to three months on 8 December 2021, following concerns over the surge in cases and the emergence of the Omicron variant (ref: <https://www.england.nhs.uk/2021/12/people-40-and-over-to-get-their-lifesaving-booster-jab-three-months-on-from-second-dose/>). Vaccine prioritisation schedules were largely determined by recommendations from the Joint Committee for Vaccine and Immunisation (JCVI) expert working group.

In most cases, BNT162b2 and mRNA-1273 vaccines were used, and administered regardless of the vaccine brand used for the first and second doses. BNT162b2 was used from 16 September 2021, and mRNA-1273 from 29 October 2021 onwards.

@@

[context]

We estimated the effectiveness of booster vaccination with BNT162b2 versus two doses only in adults in England, using the OpenSAFELY-TPP linked primary care database covering around 40% of England's population. The study is conducted in an era where the Delta variant is dominant, with follow-up curtailed shortly after the emergence of Omicron.

@@

@@

We used observational data to emulate a target trial @dagan2021 @barda2021 of the effectiveness of booster vaccination in people who had received two COVID-19 vaccine doses and were eligible for booster vaccination between `r format(as.Date(study_dates$studystart_date), "%d %B)` and `r format(as.Date(study_dates$lastvax3_date), "%d %B %Y")` inclusive.

Eligibility on each day was as follows: adults (18 years and over) who were registered at a GP practice using TPP's SystmOne clinical information system; received either two doses of BNT162b2 doses or two doses of ChAdOx1-S (mixed dosing and Moderna mRNA-1273 doses were not considered due to small numbers); not a health or social care worker, not a resident in a care or nursing home, and not medically housebound or receiving end-of-life care (due to distinct clustering of vaccine uptake and infection risk in these groups that could not be reasonably measured and controlled for); no evidence of SARS-CoV-2 infection or COVID-19 disease within the previous 90 days; not undergoing an unplanned hospital admission; information on sex, ethnicity, deprivation, and NHS region was complete.

Additionally, if the rolling weekly average count of BNT162b2 booster doses within strata defined by region, JCVI group, and week of second dose was below 50 then recruitment did not occur for that strata on that day. This helped to ensure that recruitment was restricted to those who were boosted in line with national prioritisation schedules and that the matched controls were both eligible and able to receive a booster dose at the time of potential recruitment.

@@

For each trial, the booster (treatment) group comprised people who received a BNT162b2 booster dose on that day, and the unboosted (control) group comprised the same number of randomly selected unboosted people, matched on the following characteristics: primary course vaccine type (BNT162b2 or ChAdOx1-S); date of the second vaccine dose, within 7 days; NHS region (East of England, Midlands, London, North East and Yorkshire, North West, South East, South West); any evidence of prior SARS-CoV-2 infection (positive SARS-CoV-2 test, "probable" infection documented in primary care, or COVID-19 hospitalisation); Clinical risk group, defined by JCVI (clinically extremely vulnerable, clinically at-risk, neither); age, within 3 years, and by age groups defined by JCVI. Note that the matching and exclusion criteria jointly ensure that treatment groups are also matched within JCVI priority group.

People selected as controls were not eligible to be included again as a control in a subsequent trial, but were eligible for selection into the treatment group of a subsequent trial.

@@

@@

We use Cox models, stratified by recruitment day, NHS region, and brand of second vaccine dose, to estimate hazard ratios comparing boosted with unboosted people. We estimated effectveness over all follow-up time (0-70 days), and separately in days 0-6, 7-13, 14-27, 28-41, 42-70. Models were adjusted for additional variables not used as matching factors, as described below. A robust variance estimator was used to account for participant clustering.

@@

We additionally estimated booster effectiveness in the following subgroups: primary vaccine course (both BNT162b2 or both ChAdOx1-S); age (18-64 versus 65 and over); clinically extremely vulnerable. Analyses were conducted independently in each subgroup.

Booster effectiveness is defined as one minus the hazard ratio, and expressed as a percentage.

## Covariates

The following variables were included as covariates: sex (male or female); English Index of Multiple Deprivation (IMD, grouped by quintile); ethnicity (Black, Mixed, South Asian, White, Other, as per the UK census); morbidity count (diabetes, BMI > 40, chronic heart disease, chronic kidney disease, chronic liver disease, chronic respiratory disease or severe asthma, chronic neurological disease); learning disabilities; severe mental illness; the number of SARS-CoV-2 tests in the 6 months prior to the study start date; elective in-hospital episode at the time of recruitment.

@@

Total follow-up and incidence rates for each outcome, by treatment group. Cumulative incidences are provided in Supplementary Figure S2.

```{r table2, echo=FALSE, warning=FALSE, message=FALSE}
incidences %>%
  filter(
    fup_period == "All",
    outcome %in% c("postest", "covidadmitted", "coviddeath"),
    treatment == "pfizer"
  ) %>%
  left_join(
    matchsummary %>% select(treatment, n_eligible, prop_matched_eligible),
    by=c("treatment")
  ) %>%
  mutate(
    treatment_descr = paste0(
      treatment_descr, " - ",
      label_number(1, big.mark=",")(n_eligible), "eligible, ",
      label_number(1, big.mark=",")(n_1), "matched (",
      label_percent(0.1)(prop_matched_eligible), ")."
    )
  ) %>%
  select(
    treatment_descr, 
    outcome_descr,
    q_1,
    rate_1,
    q_0, 
    rate_0,
    rrECI
  ) %>%
  mutate(
    rate_1=if_else(rate_1<0.001, "<0.001", label_number(0.001)(rate_1)),
    rate_0=if_else(rate_0<0.001, "<0.001", label_number(0.001)(rate_0)),
  ) %>%
  gt(
    groupname_col = c("treatment_descr"),
  ) %>%
  cols_label(
    #treatment_descr = "Treatment",
    outcome_descr = "Outcome",

@@

@@

<!-- `r filter(incidences, treatment=="pfizer", outcome=="covidattendance", fup_period=="all") %>% pull(events)` COVID-19 A&E attendances, `r filter(incidences, treatment=="pfizer", outcome=="covidcc", fup_period=="all") %>% pull(events)` COVID-19 ciritical care admissions, -->

In the 14-20 days period of follow-up, the estimated effectiveness (95% CI) comparing a BNT162b2 booster dose to two doses only, was `r filter(hazards, subgroup=="none", outcome=="postest", treatment=="pfizer", model==3, term_left==14) %>% pull(ve_display)` for documented infection, `r filter(hazards, subgroup=="none", outcome=="covidadmitted", treatment=="pfizer", model==3, term_left==14) %>% pull(ve_display)` for COVID-19 hospital admission, and `r filter(hazards, subgroup=="none", outcome=="coviddeath", treatment=="pfizer", model==3, term_left==14) %>% pull(ve_display)` for COVID-19 death (Figure 2).

For days 42-70, estimated effectiveness (95% CI) comparing a BNT162b2 booster dose to two doses only, was `r filter(hazards, subgroup=="none", outcome=="postest", treatment=="pfizer", model==3, term_left==42) %>% pull(ve_display)` for documented infection, `r filter(hazards, subgroup=="none", outcome=="covidadmitted", treatment=="pfizer", model==3, term_left==42) %>% pull(ve_display)` for COVID-19 hospital admission, and `r filter(hazards, subgroup=="none", outcome=="coviddeath", treatment=="pfizer", model==3, term_left==42) %>% pull(ve_display)` for COVID-19 death (Figure 2), suggesting a potential waning of booster effectiveness with time. 
<!-- `r filter(hazards, subgroup=="none", outcome=="covidattendance", treatment=="pfizer", model==3, term_left==14) %>% pull(ve_display)` for COVID-19 A&E attendances, `r filter(hazards, subgroup=="none", outcome=="covidcc", treatment=="pfizer", model==3, term_left==14) %>% pull(ve_display)` for COVID-19 critical care admission,  -->

<!-- For mRNA-1273, the corresponding estimated effectiveness were `r filter(hazards, subgroup=="none", outcome=="postest", treatment=="moderna", model==3, term_left==14) %>% pull(ve_display)`, for COVID-19 A&E attendances was `r filter(hazards, subgroup=="none", outcome=="covidattendance", treatment=="moderna", model==3, term_left==14) %>% pull(ve_display)`, for COVID-19 hospital admission was `r filter(hazards, subgroup=="none", outcome=="covidadmitted", treatment=="moderna", term_left==14) %>% pull(ve_display)`, for COVID-19 critical care admission was `r filter(hazards, subgroup=="none", outcome=="covidcc", treatment=="moderna", model==3, term_left==14) %>% pull(ve_display)`, and for COVID-19 death was `r filter(hazards, subgroup=="none", outcome=="coviddeath", treatment=="moderna", model==3, term_left==14) %>% pull(ve_display)`.  -->

For the same period, the estimated effectiveness (95% CI) comparing a BNT162b2 booster dose to two doses only separately for 2× BNT162b2 and 2× ChAdOx1-S primary course recipients respectively were `r filter(hazards, outcome=="postest", treatment=="pfizer", subgroup_level=="pfizer-pfizer", model==3, term_left==14) %>% pull(ve_display)` and `r filter(hazards, outcome=="postest", treatment=="pfizer", subgroup_level=="az-az", model==3, term_left==14) %>% pull(ve_display)` for documented infection, `r filter(hazards, outcome=="covidadmitted", treatment=="pfizer", subgroup_level=="pfizer-pfizer", model==3, term_left==14) %>% pull(ve_display)` and `r filter(hazards, outcome=="covidadmitted", treatment=="pfizer", subgroup_level=="az-az", model==3, term_left==14) %>% pull(ve_display)` for COVID-19 hospital admission, and `r filter(hazards, outcome=="coviddeath", treatment=="pfizer", subgroup_level=="pfizer-pfizer", model==3, term_left==14) %>% pull(ve_display)` and `r filter(hazards, outcome=="coviddeath", treatment=="pfizer", subgroup_level=="az-az", model==3, term_left==14) %>% pull(ve_display)` for COVID-19 death (Figure 2).

@@

>>>>

---
title: "Effectiveness of BNT162b2 booster doses in England: a observational study in OpenSAFELY-TPP"
output:
  html_document:
    keep_md: yes
    self_contained: yes
  word_document: default
  bookdown::html_document2:
    number_sections: no
    toc: no
  pdf_document:
    toc: no
bibliography: references.bib
csl: vancouver.csl
link-citations: yes
zotero: yes
header-includes:
- \usepackage{float} 
- \floatplacement{figure}{H}
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \captionsetup[table]{labelformat=empty}
---

@@

cohortsummary <- read_csv(fs::path(output_dir_os, "cohortsummary.csv"))
matchsummary <- read_csv(fs::path(output_dir_os, "matchsummary.csv"))
matchsummary_treated <- read_csv(fs::path(output_dir_os, "matchsummary_treated.csv")) %>% filter(treated == 1L)
matchsummary_control <- read_csv(fs::path(output_dir_os, "matchsummary_treated.csv")) %>% filter(treated == 0L)
matchflowchart <- read_csv(fs::path(output_dir_os, "matchflowchart.csv"))
matchcoverage <- read_csv(fs::path(output_dir_os, "matchcoverage.csv"))
incidences <- read_csv(fs::path(output_dir_os, "none", "incidence_all.csv")) %>% mutate(events=events_1+events_0)

@@

  outcomes <- 
    c(
      "postest",
      #"covidemergency",
      "covidadmitted",
      #"covidcc",
      "coviddeath",
      "noncoviddeath"
    ) %>% 
    set_names(., .)

km <- 
  bind_rows(
    read_csv(fs::path(output_dir_os, "none", "km.csv")),
    read_csv(fs::path(output_dir_os, "vax12_type", "km.csv")),
    read_csv(fs::path(output_dir_os, "cev", "km.csv")) %>% mutate(subgroup_level=as.character(subgroup_level)),
    read_csv(fs::path(output_dir_os, "age65plus", "km.csv")) %>% mutate(subgroup_level=as.character(subgroup_level))
  ) %>%
  filter(
    outcome %in% outcomes
  ) %>%
  mutate(
    subgroup_descr = fct_inorder(subgroup_descr),
    subgroup_level = fct_inorder(subgroup_level),
    subgroup_level_descr = fct_inorder(fct_explicit_na(subgroup_level_descr, "All")),
    treatment_descr = fct_inorder(treatment_descr),
    outcome_descr = fct_inorder(outcome_descr),
    outcome_descr_wrap = fct_relabel(outcome_descr, ~str_wrap(., width=15)),
    #model_name = fct_inorder(str_wrap(model_name, 15)),
  )
# 
# cmlinc <- read_csv(fs::path(output_dir_os, "cmlinc.csv")) %>%
#   mutate(
#     treatment_descr = fct_inorder(treatment_descr),
#     outcome_descr = fct_inorder(outcome_descr),
#     model_name = fct_inorder(str_wrap(model_name, 15)),
#     inc_CI_to = paste0(label_number(0.1, scale=1000)(1-survival.ul), " to ", label_number(0.1, scale=1000)(1-survival.ll))
#   )

@@

hazards <- 
  bind_rows(
    read_csv(fs::path(output_dir_os, "none", "effects.csv")),
    read_csv(fs::path(output_dir_os, "vax12_type", "effects.csv")),
    read_csv(fs::path(output_dir_os, "cev", "effects.csv")) %>% mutate(subgroup_level=as.character(subgroup_level)),
    read_csv(fs::path(output_dir_os, "age65plus", "effects.csv"))  %>% mutate(subgroup_level=as.character(subgroup_level)),
  ) %>%
  filter(
    outcome %in% outcomes
  ) %>%
  mutate(

@@

overallhazards <- 
  bind_rows(
    read_csv(fs::path(output_dir_os, "none", "overalleffects.csv")),
    read_csv(fs::path(output_dir_os, "vax12_type", "overalleffects.csv")),
    read_csv(fs::path(output_dir_os, "cev", "overalleffects.csv")) %>% mutate(subgroup_level=as.character(subgroup_level)),
    read_csv(fs::path(output_dir_os, "age65plus", "overalleffects.csv")) %>% mutate(subgroup_level=as.character(subgroup_level)),
  ) %>%
  filter(
    outcome %in% outcomes
  ) %>%
  mutate(

@@

meta2hazards <- 
  # derive directly from hazards as meta2effects table needs to be re-run
  bind_rows(
    read_csv(fs::path(output_dir_os, "none", "effects.csv")),
    read_csv(fs::path(output_dir_os, "vax12_type", "effects.csv")),
    read_csv(fs::path(output_dir_os, "cev", "effects.csv")) %>% mutate(subgroup_level=as.character(subgroup_level)),
    read_csv(fs::path(output_dir_os, "age65plus", "effects.csv"))  %>% mutate(subgroup_level=as.character(subgroup_level)),
  ) %>%
  filter(
    outcome %in% outcomes
  ) %>%
  mutate(
    period = (term_right>28) + 1,
    subgroup_descr = fct_inorder(subgroup_descr),
    subgroup_level = fct_inorder(subgroup_level),
    subgroup_level_descr = fct_inorder(fct_explicit_na(subgroup_level_descr, "All")),
    treatment_descr = fct_inorder(treatment_descr),
    outcome_descr = fct_inorder(outcome_descr),
    outcome_descr_wrap = fct_relabel(outcome_descr, ~str_wrap(., width=20)),
  ) %>%
  group_by(
    treatment, treatment_descr, outcome, outcome_descr, outcome_descr_wrap, subgroup, subgroup_variable, subgroup_descr, subgroup_level, subgroup_level_descr, model, model_descr, period, 
  ) %>%
  summarise(
    term_left = min(term_left),
    term_right = max(term_right),
    term_midpoint = (term_left+term_right)/2,
    estimate = weighted.mean(estimate, robust.se^-2),
    std.error = sqrt(1/sum(std.error^-2)),
    robust.se = sqrt(1/sum(robust.se^-2)),
    statistic = estimate/robust.se,
    p.value = 2 * pmin(pnorm(statistic), pnorm(-statistic)),
    conf.low = estimate + qnorm(0.025)*robust.se,
    conf.high = estimate + qnorm(0.975)*robust.se,
    hr = exp(estimate),
    hr.ll = exp(conf.low),
    hr.ul = exp(conf.high),
    ve = 1-hr,
    ve.ll = 1-hr.ul,
    ve.ul = 1-hr.ll
  ) %>%
  ungroup() %>%
  mutate(
    hr_display = paste0(label_number(accuracy=0.01)(hr), " (", label_number(accuracy=0.01)(hr.ll), "-", label_number(accuracy=0.01)(hr.ul), ")"),
    ve_display = paste0(label_percent(accuracy=0.1)(ve), " (", label_percent(accuracy=0.1, suffix="")(ve.ll), "-", label_percent(accuracy=0.1, suffix="")(ve.ul), ")"),
  )

# meta2hazards <- 
#   bind_rows(
#     read_csv(fs::path(output_dir_os, "none", "meta2effects.csv")),
#     read_csv(fs::path(output_dir_os, "vax12_type", "meta2effects.csv")),
#     read_csv(fs::path(output_dir_os, "cev", "meta2effects.csv")) %>% mutate(subgroup_level=as.character(subgroup_level)),
#     read_csv(fs::path(output_dir_os, "age65plus", "meta2effects.csv")) %>% mutate(subgroup_level=as.character(subgroup_level)),
#   ) %>%
#   filter(
#     outcome %in% outcomes
#   ) %>%
#   mutate(
#     subgroup_descr = fct_inorder(subgroup_descr),
#     subgroup_level = fct_inorder(subgroup_level),
#     subgroup_level_descr = fct_inorder(fct_explicit_na(subgroup_level_descr, "All")),
#     treatment_descr = fct_inorder(treatment_descr),
#     outcome_descr = fct_inorder(outcome_descr),
#     outcome_descr_wrap = fct_relabel(outcome_descr, ~str_wrap(., width=20)),
#     #model_name = fct_inorder(str_wrap(model_name, 15)),
#     hr_display = paste0(label_number(accuracy=0.01)(hr), " (", label_number(accuracy=0.01)(hr.ll), "-", label_number(accuracy=0.01)(hr.ul), ")"),
#     ve_display = paste0(label_percent(accuracy=0.1)(ve), " (", label_percent(accuracy=0.1, suffix="")(ve.ll), "-", label_percent(accuracy=0.1, suffix="")(ve.ul), ")"),
#   )

@@

With the approval of NHS England we used the OpenSAFELY-TPP database, covering 40% of GP practices in England and linked to national coronavirus surveillance, hospital episodes, and death registry data, to estimate the effectiveness of BNT162b2 boosting in adults eligible for boosting between `r format(as.Date(study_dates$studystart_date), "%d %B %Y")` and `r format(as.Date(study_dates$lastvax3_date), "%d %B %Y")`. Each booster recipient was matched with an unboosted control on factors relating to booster priority status and prior immunisation. Primary outcomes were documented SARS-CoV-2 infection, COVID-19 hospital admission, and COVID-19 death up to 10 weeks after recruitment. Hazard ratios were estimated using Cox models adjusting for additional covariates that were not used as matching factors. Follow-up spanned a period in which the Delta variant was predominant.

@@

We matched `r label_number(1, big.mark=",")(filter(matchsummary, treatment=="pfizer") %>% pull(n_matched))` BNT162b2 booster recipients to unboosted controls. During days 15-28 after the booster, the estimated hazard ratio (95% CI) comparing a BNT162b2 booster dose to two doses only was `r filter(hazards, subgroup=="none", outcome=="postest", treatment=="pfizer", model==3, term_right==28) %>% pull(ve_display)` for documented SARS-CoV-2 infection, `r filter(hazards, subgroup=="none", outcome=="covidadmitted", treatment=="pfizer", model==3, term_right==28) %>% pull(ve_display)` for COVID-19 hospital admission, and `r filter(hazards, subgroup=="none", outcome=="coviddeath", treatment=="pfizer", model==3, term_right==28) %>% pull(ve_display)` for COVID-19 death.

@@

<!---# *Participants* `r label_number(1, big.mark=",")(filter(matchflowchart, crit=="c4") %>% pull(n))` people with a complete primary COVID-19 vaccine course, registered with a GP practice using the TPP SystmOne clinical information system in England. --->

@@

@@

The national COVID-19 vaccination programme in England began a new phase in September 2021 with administration of the first booster COVID-19 vaccine doses in those who had already received their primary vaccination course (a completed first and second dose) (ref <https://www.england.nhs.uk/2021/09/nhs-begins-covid-19-booster-vaccination-campaign/>). Initially offered to those at highest risk of severe disease, eligibility was progressively extended and by 15 December 2021 every adult in England was eligible (<https://www.england.nhs.uk/2021/12/nhs-booster-bookings-open-to-every-eligible-adult/>). Booster doses were initially available no earlier than six months after dose two, but this was reduced to three months on 8 December 2021, following concerns over the surge in cases and the emergence of the Omicron variant (ref: <https://www.england.nhs.uk/2021/12/people-40-and-over-to-get-their-lifesaving-booster-jab-three-months-on-from-second-dose/>). Vaccine prioritisation schedules were guided by recommendations from the Joint Committee for Vaccine and Immunisation (JCVI) expert working group.

@@

<!-- In most cases, BNT162b2 and half-dose mRNA-1273 vaccines were used for boosting, and administered regardless of the first and second dose brand. BNT162b2 was used from 16 September 2021, and mRNA-1273 from 29 October 2021 onwards. This provides an opportunity to investigate booster effectiveness of two different brands administered in overlapping periods and community contexts, as well as in subgroups defined by prior immunological and other clinical factors.-->

We set out to emulate a target trial estimating the effectiveness of booster vaccination with BNT162b2 against COVID-19 outcomes in adults who had already received two COVID-19 vaccine doses. We used the OpenSAFELY-TPP linked primary care database covering around 40% of English residents. Recruitment covers an era where the Delta variant is dominant, with follow-up curtailed shortly after the emergence of Omicron.

@@

## Study design

We used observational data to emulate a target trial of the effectiveness of booster vaccination @dagan2021 @barda2021, by comparing people who had received a BNT162b2 booster dose with matched people who had not.

@@

We recruited adults who were eligible for booster vaccination between `r format(as.Date(study_dates$studystart_date), "%d %B")` and `r format(as.Date(study_dates$lastvax3_date), "%d %B %Y")` inclusive. Eligibility on each day was as follows: adults (18 years and over) who were registered at a GP practice using TPP's SystmOne clinical information system; received either two doses of BNT162b2 doses or two doses of ChAdOx1-S (mixed dosing and Moderna mRNA-1273 doses were not considered due to small numbers); not a health or social care worker, not a resident in a care or nursing home, and not medically housebound or receiving end-of-life care (due to distinct clustering of vaccine uptake and infection risk in these groups that could not be reasonably measured and controlled for); no evidence of SARS-CoV-2 infection or COVID-19 disease within the previous 90 days; not undergoing an unplanned hospital admission; information on sex, ethnicity, deprivation, and NHS region was complete.

Additionally, if the rolling weekly average count of BNT162b2 booster doses within strata defined by region, JCVI group, and the week of second dose was below 50 then recruitment did not occur for that strata on that day. This helped to ensure that recruitment was restricted to those who were boosted in line with national prioritisation schedules and that the matched controls were both eligible and able to receive a booster dose at the time of potential recruitment.

@@

On each day of recruitment, each eligible person receiving a booster dose on that day was recruited to the treatment group and matched with an unboosted control person who had received only two doses. Pairs were matched on the following characteristics: primary course vaccine brand (2× BNT162b2 or 2× ChAdOx1-S); date of the second vaccine dose, within 7 days; NHS region (East of England, Midlands, London, North East and Yorkshire, North West, South East, South West); any evidence of prior SARS-CoV-2 infection (positive SARS-CoV-2 test, "probable" infection documented in primary care, or COVID-19 hospitalisation); clinical risk groups used for prioritisation (clinically extremely vulnerable, clinically at-risk, neither); age, within 3 years, and by age groups used for prioritisation. The matching and exclusion criteria jointly ensure that treatment groups are also matched within JCVI priority group.

People selected as controls were not eligible to be included again as a control in a subsequent trial, but were eligible for selection into the treatment group of a subsequent trial. Any unmatched boosted people were dropped.

@@

## Additional variables

The following additional variables were included as covariates in the Cox models (see below) or else to describe the treatment groups: sex (male or female); English Index of Multiple Deprivation (IMD, grouped by quintile); ethnicity (Black, Mixed, South Asian, White, Other, as per the UK census); morbidity count (diabetes, BMI > 40kg/m^2^, chronic heart disease, chronic kidney disease, chronic liver disease, chronic respiratory disease or severe asthma, chronic neurological disease); learning disabilities; severe mental illness; the number of SARS-CoV-2 tests in the 6 months prior to the study start date; elective in-hospital episode at the time of recruitment.

@@

We use Cox models, stratified by recruitment day, NHS region, and brand of second vaccine dose, to estimate hazard ratios comparing boosted with unboosted people. We estimated effectveness over all follow-up time (1-70 days), and separately in days 1-7, 8-14, 15-28, 29-42, 43-70. Models were adjusted for additional variables not used as matching factors, as described below. A robust variance estimator was used to account for participant clustering.

@@

We also estimated booster effectiveness in the following subgroups: primary vaccine course (both BNT162b2 or both ChAdOx1-S); age (18-64 or 65 and over); clinically extremely vulnerable. Analyses were conducted separately in each subgroup.

Booster effectiveness is defined as one minus the hazard ratio, expressed as a percentage.

@@

Total follow-up and incidence rates for each outcome, by treatment group. The cumulative incidence of each outcome is provided in Supplementary Figure S2.

```{r table2, echo=FALSE, warning=FALSE, message=FALSE}
incidences %>%
  filter(
    fup_period == "All",
    outcome %in% c("postest", "covidadmitted", "coviddeath"),
    treatment == "pfizer"
  ) %>%
  left_join(
    matchsummary %>% select(treatment, n_eligible, prop_matched_eligible),
    by=c("treatment")
  ) %>%
  mutate(
    treatment_descr = paste0(
      #treatment_descr, " - ",
      label_number(1, big.mark=",")(n_eligible), " eligible, ",
      label_number(1, big.mark=",")(n_1), " matched (",
      label_percent(0.1)(prop_matched_eligible), ")."
    )
  ) %>%
  select(
    treatment_descr, 
    outcome_descr,
    q_1,
    rate_1,
    q_0, 
    rate_0,
    rrECI
  ) %>%
  mutate(
    rate_1=if_else(rate_1<0.001, "<0.001", label_number(0.001)(rate_1)),
    rate_0=if_else(rate_0<0.001, "<0.001", label_number(0.001)(rate_0)),
  ) %>%
  gt(
   groupname_col = c("treatment_descr"),
  ) %>%
  cols_label(
    #treatment_descr = "Treatment",
    outcome_descr = "Outcome",

@@

### Main analysis

@@

The estimated effectiveness (95% CI) comparing a BNT162b2 booster dose to two doses only over all follow-up time was `r filter(overallhazards, subgroup=="none", outcome=="postest", treatment=="pfizer", model==3) %>% pull(ve_display)` for documented infection, `r filter(overallhazards, subgroup=="none", outcome=="covidadmitted", treatment=="pfizer", model==3) %>% pull(ve_display)` for COVID-19 hospital admission, and `r filter(overallhazards, subgroup=="none", outcome=="coviddeath", treatment=="pfizer", model==3) %>% pull(ve_display)` for COVID-19 death. Effectiveness before and after 28 days since boosting is also estimated (Figure 2, Table 3), as well as for narrower periods (Supplementary Table S3).

<!-- In the first 28 days of follow-up the estimated effectiveness (95% CI) comparing a BNT162b2 booster dose to two doses only, was `r filter(meta2hazards, subgroup=="none", outcome=="postest", treatment=="pfizer", model==3, term_right==28) %>% pull(ve_display)` for documented infection, `r filter(meta2hazards, subgroup=="none", outcome=="covidadmitted", treatment=="pfizer", model==3, term_right==28) %>% pull(ve_display)` for COVID-19 hospital admission, and `r filter(meta2hazards, subgroup=="none", outcome=="coviddeath", treatment=="pfizer", model==3, term_right==28) %>% pull(ve_display)` for COVID-19 death (Figure 2). -->

<!-- For days 28-70, estimated effectiveness (95% CI) comparing a BNT162b2 booster dose to two doses only, was `r filter(meta2hazards, subgroup=="none", outcome=="postest", treatment=="pfizer", model==3, term_right==70) %>% pull(ve_display)` for documented infection, `r filter(meta2hazards, subgroup=="none", outcome=="covidadmitted", treatment=="pfizer", model==3, term_right==70) %>% pull(ve_display)` for COVID-19 hospital admission, and `r filter(meta2hazards, subgroup=="none", outcome=="coviddeath", treatment=="pfizer", model==3, term_right==70) %>% pull(ve_display)` for COVID-19 death (Figure 2), suggesting a potential waning of booster effectiveness with time. -->

### Subgroup analyses

The estimated effectiveness (95% CI) for 2× BNT162b2 and 2× ChAdOx1-S primary course recipients respectively were `r filter(overallhazards, outcome=="postest", treatment=="pfizer", subgroup_level=="pfizer-pfizer", model==3) %>% pull(ve_display)` and `r filter(overallhazards, outcome=="postest", treatment=="pfizer", subgroup_level=="az-az", model==3) %>% pull(ve_display)` for documented infection, `r filter(overallhazards, outcome=="covidadmitted", treatment=="pfizer", subgroup_level=="pfizer-pfizer", model==3) %>% pull(ve_display)` and `r filter(overallhazards, outcome=="covidadmitted", treatment=="pfizer", subgroup_level=="az-az", model==3) %>% pull(ve_display)` for COVID-19 hospital admission, and `r filter(overallhazards, outcome=="coviddeath", treatment=="pfizer", subgroup_level=="pfizer-pfizer", model==3) %>% pull(ve_display)` and `r filter(overallhazards, outcome=="coviddeath", treatment=="pfizer", subgroup_level=="az-az", model==3) %>% pull(ve_display)` for COVID-19 death (Figure 2).

Period-specific estimates are also provided.

@@

## Table 3: Estimated booster effectiveness

Main and subgroup analyses
```{r table_effects, echo=FALSE, message=FALSE, warning=FALSE, out.extra='', fig.height=7.5, out.width='90%', results='asis'}
